Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-05-29 sources/bmc_template_submission/bmc_article.tex
last tiny edits

As a tool for automatizing phylogenetic reconstruction from molecular databases,
Physcraper presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data available. Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input phylogeny, Physcraper estimates all
the relationships anew in the context of the new data.
The tool PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has produced
few published phylogenetic analyses since its publication.
Several other existing tools create full phylogenies \textit{de novo} by mining of molecular
databases \cite{antonelli2017toward, sanderson2008phylota, smith2019pyphlawd,
bennett2018phylotar, pearse2013phylogenerator}. In particular, Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega}, have been cited and used abundantly.
Physcraper adds to this automated database mining concept by incorporating prior
phylogenetic work and existing taxonomic domain knowledge on appropriate markers
and alignment construction, to update existing phylogenetic knowledge.
This decreases error (requiring less manual downstream processing) and
eases comparison with previous phylogenetic hypotheses.
Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.

We encourage users to look at the output files containing information about the
filtered sequences, and potentially modify configuration parameters such as the
sequence length cutoff parameter, based on said information.
Physcraper's default filtering parameters are arbitrary, but we hope that by making the process
of locating homologous sequences online reproducible, and tracking what filters
are used, we make it easier for researchers to delve into the effect that different
filtering choices have on their inferences.
This is in contrast to ``manual" searches for taxa, where similarly arbitrary filters
are applied, but are difficult to trace.
It has been shown \cite{huang_what_2009} that the effect of missing data in alignments can
be enigmatic, and interact with the true phylogenetic relationships for the dataset at hand.
There is not currently strong support in the literature for any particular filtering cutoff
value, and rather than prescribe specific approaches, we encourage users to explore
the effects of different parameter values on resulting phylogenetic hypotheses.
In addition, by providing the output files at each step of the analysis,
it is straightforward to assess how changing filtering cutoffs and software choices might drive differences in phylogenetic inference.
By gathering the DNA sequences, and making the unaligned files easily accessible and reusable,
Physcraper facilitates the exploration of alternative aligning tools to determine their effect on phylogenetic inferences.
Once sequences are aligned, users can apply practically any phylogenetic
software and compare results using Physcraper tools in a reproducible framework.

@@

\section*{Conclusions}
Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy, applications and education.
Scientific understanding is constantly challenged and reframed by new data and analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently contextualize new knowledge in the light of previous understanding,
showcasing the utility and importance of good data sharing practices and open science for phylogenetics, biology and science.

>>>>

As a tool for automatizing phylogenetic reconstruction from molecular databases,
Physcraper presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public genetic data available. Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input phylogeny, Physcraper estimates all
the relationships anew in the context of the new data.
The tool PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has produced
few published phylogenetic analyses since its publication.
Several other existing tools create full phylogenies \textit{de novo} by mining of molecular
databases \cite{antonelli2017toward, sanderson2008phylota, smith2019pyphlawd,
bennett2018phylotar, pearse2013phylogenerator}. In particular, Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega}, have been cited and used abundantly.
Physcraper adds to this automated database mining concept by incorporating prior
phylogenetic work and existing taxonomic domain knowledge on appropriate markers
and alignment construction, to update existing phylogenetic knowledge.
This decreases error (requiring less manual downstream processing) and
eases comparison with previous phylogenetic hypotheses.
Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.

We encourage users to look at Physcraper's output files containing information about the
filtered sequences, and use this results to potentially modify configuration parameters such as the
sequence length cutoff parameter in subsequent Physcraper runs.
Physcraper's default filtering parameters are arbitrary, but we hope that by making the process
of locating homologous sequences online reproducible, and tracking what filters
are used, we make it easier for researchers to delve into the effect that different
filtering choices have on their inferences.
This is in contrast to ``manual" searches for DNA sequences in molecular databases, where similarly arbitrary filters
are applied, but are difficult to trace.
It has been shown \cite{huang_what_2009} that the effect of missing data in alignments can
be enigmatic, and interact with the true phylogenetic relationships for the dataset at hand.
There is not currently strong support in the literature for any particular filtering cutoff
value, and rather than prescribe specific approaches, we encourage users to explore
the effects of different parameter values on resulting phylogenetic hypotheses.
In addition, by providing the output files at each step of the analysis,
it is straightforward to assess how changing filtering cutoffs and software choices might drive differences in phylogenetic inference.
By gathering the DNA sequences, and making the unaligned files easily accessible and reusable,
Physcraper facilitates the exploration of alternative aligning tools.
Once sequences are aligned, users can apply practically any phylogenetic
software and compare results using Physcraper tools in a reproducible framework.

@@

\section*{Conclusions}
Data repositories hold more information than meets the eye.
Beyond the main data, they are rich sources of metadata that can be leveraged
for the advantage of all areas of biology as well as the advancement of scientific
policy, applications and education.
Scientific understanding is constantly challenged and reframed by new data and analyses.
Physcraper provides a framework for reproducible phylogenetics that has the
potential to consistently contextualize new knowledge in the light of previous understanding,
showcasing the utility and importance of good data sharing practices and open science for the advancement of phylogenetics, biology and research.
